A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations (MICE)

### 2019-12-04 Simulation/Writeup/NoteConvergence.tex
Finally integrated comments, also created nicer graph and worked on better flow in introduction section.

Post-processing of convergence diagnostics:

@@

Figure \ref{resultssim} shows the results over 1000 simulations. A subset of simulation conditions is presented in Table \ref{subsetresults}.

\begin{figure}[h]
  \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
        \includegraphics{Figures/8plotresults.pdf}
  }      
  \caption{Simulation and convergence diagnostics over 1000 MCMC simulations.}
    \label{resultssim}
\end{figure}

@@

%% --------------------------------
\subsection{Simulation diagnostics}
Bias. After 1 or 2 iterations bias fluctuates within a narrow range of the average bias across all simulation conditions. Average bias is not zero because there is only one incomplete dataset to be imputed. Bias is a sample effect.

Empirical standard error (SE).

Confidence interval width. Does not show a trend across the simulation conditions. [Is related to empirical SE??]

Coverage rate. Is more or less stable from two iterations upwards. But higher percentages then the expected 95\% \cite{neym43}. This is due to pooled SE??

@@

$\widehat{R}$. Steep drop down up-to iteration 20-30, then stable. Conventional threshold $\widehat{R} < 1.2$ reached after three iterations, $\widehat{R} < 1.1$ after four iterations, $\widehat{R} < 1.01$ after 38 iterations (see online appendix).

Auto-correlation. Auto-correlation is dangerous when positive. These auto-correlations are negative. Still, we want the iterations to be stable and independent. Default maxit is five iterations now, should this be different? Are the 'waves' most pronounced at iteration 5? But why the dip??

% latex table generated in R 3.6.1 by xtable 1.8-4 package
% Fri Nov 29 11:38:27 2019
\begin{table}[ht]
\centering
\caption{Simulation and convergence diagnostics over 1000 MCMC simulations.} 
\label{results}
\begin{tabular}{lrrrrrrr}
  \hline
It. & Bias & CI width & Cov. rate & $\widehat{R}$ & $\widehat{R}$ & Auto-corr. & Auto-corr. \\ 
  \hline
   1 & -0.128 & 0.922 & 0.930 & NA & NA & NA & NA \\ 
     2 & -0.002 & 0.915 & 0.940 & 1.686 & 1.709 & -0.500 & -0.500 \\ 
     3 & 0.012 & 0.923 & 0.960 & 1.317 & 1.326 & -0.661 & -0.659 \\ 
     4 & 0.019 & 0.892 & 0.970 & 1.461 & 1.464 & -0.735 & -0.746 \\ 
     5 & 0.015 & 0.906 & 0.910 & 1.486 & 1.472 & -0.706 & -0.705 \\ 
     6 & 0.010 & 0.928 & 0.950 & 1.252 & 1.261 & -0.651 & -0.667 \\ 
     7 & 0.006 & 0.908 & 0.910 & 1.259 & 1.267 & -0.619 & -0.584 \\ 
     8 & 0.018 & 0.920 & 0.940 & 1.178 & 1.181 & -0.516 & -0.549 \\ 
     9 & 0.011 & 0.994 & 0.940 & 1.190 & 1.186 & -0.506 & -0.488 \\ 
    10 & 0.020 & 0.940 & 0.900 & 1.142 & 1.145 & -0.448 & -0.420 \\ 
    15 & 0.009 & 0.917 & 0.960 & 1.098 & 1.109 & -0.256 & -0.289 \\ 
    20 & 0.020 & 0.848 & 0.930 & 1.069 & 1.065 & -0.179 & -0.214 \\ 
    25 & 0.026 & 0.897 & 0.970 & 1.054 & 1.054 & -0.146 & -0.131 \\ 
    50 & 0.008 & 0.946 & 0.970 & 1.026 & 1.026 & -0.029 & -0.096 \\ 
   100 & 0.011 & 0.967 & 0.930 & 1.013 & 1.013 & 0.054 & -0.020 \\ 
   \hline
\end{tabular}
\end{table}

From the simulation diagnostics it appears that as little as three iterations is sufficient to draw valid inference. Convergence diagnostics $\widehat{R}$ and auto-correlation, however, indicate 20-30 iterations. Apparently, they are not perfect for MI data [move interpretation to Discussion!].

[Add simulation time (computational cost) to table? Add empirical SE of all diagnostics? Sample some random simulations and check distribution and plausibility of imputations? Use chain variance as scalar of interest for rhat and autocorrelation?]

%% -- Manuscript ---------------------------------------------------------------

%% -- Illustrations ------------------------------------------------------------

@@

This note illustrates that conventional convergence diagnostics behave differently on MI data as compared to the output of 'regular' MCMC algorithms. The recommended threshold for $\widehat{R}$ is too stringent for MI data.

Moreover, $\widehat{R}$ could theoretically not be smaller than one, yet it happened several times in this study (see online appendix XYZ). How could $\widehat{R}$ smaller than 1 occur? The number of simulations is smaller than in `regular' MCMC processes [explain that less iterations is an advantage of MI, not a disadvantage compared to MCMC]. Therefore, the `$(n-1/n)$' [add equation number] correction factor can influence the estimated potential scale reduction factor. This downwards bias is in the opposite direction than expected: ``The mixture-of-sequences variance, V, should stabilize as a function of n. (Before convergence, we expect $\sigma^2$ to decrease with n, only increasing if the sequences explore a new area of parameter space, which would imply that the original sequences were not overdispersed for the particular scalar summary being monitored.)'' \cite[p~438]{broo98}.

@@

Future research: We would like to have convergence measures for multivariable statistics (scalars?) of interest. This is, however, dependent of the complete data model. The eigenvector decomposition method proposed by McKay (?) should be implemented. I could not find any resources to apply this method and it is outside the scope of this thesis to investigate how this approach could be implemented.

[final par is too negative, explain scope in intro instead.]

>>>>

@@

Figure \ref{fig:results} shows the results over 1000 simulations. A subset of simulation conditions is presented in Table \ref{tab:results}.

\includegraphics{NoteConvergence-003}

\textbf{Try to put the chain means and vars together in one panel}

% \begin{figure}[h]
%   \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
%         \includegraphics{Figures/results.pdf}
%   }
%   \caption{Simulation and convergence diagnostics over 1000 MCMC simulations.}
%     \label{resultssim}
% \end{figure}

@@

% latex table generated in R 3.6.1 by xtable 1.8-4 package
% Wed Dec 04 18:16:11 2019
\begin{table}[ht]
\centering
\caption{Simulation and convergence diagnostics over 1000 MCMC simulations.} 
\label{tab:results}
\begin{tabular}{lrrrrrrr}
  \hline
It. & Bias & CI width & Cov. rate & $\widehat{R}_{mean}$ & $\widehat{R}_{var}$ & $AC_{mean}$ & $AC_{var}$ \\ 
  \hline
   1 & -0.137 & 0.954 & 0.932 & NA & NA & NA & NA \\ 
     2 & -0.006 & 0.932 & 0.953 & 1.650 & 1.632 & -0.500 & -0.500 \\ 
     3 & 0.002 & 0.929 & 0.944 & 1.314 & 1.306 & -0.660 & -0.659 \\ 
     4 & 0.003 & 0.933 & 0.957 & 1.461 & 1.457 & -0.733 & -0.735 \\ 
     5 & 0.004 & 0.935 & 0.954 & 1.475 & 1.472 & -0.705 & -0.706 \\ 
     6 & 0.001 & 0.934 & 0.956 & 1.258 & 1.256 & -0.656 & -0.646 \\ 
     7 & 0.002 & 0.930 & 0.948 & 1.265 & 1.269 & -0.591 & -0.585 \\ 
     8 & 0.004 & 0.930 & 0.954 & 1.181 & 1.178 & -0.495 & -0.516 \\ 
     9 & 0.003 & 0.931 & 0.956 & 1.187 & 1.187 & -0.442 & -0.459 \\ 
    10 & 0.003 & 0.952 & 0.943 & 1.141 & 1.140 & -0.403 & -0.423 \\ 
    15 & 0.002 & 0.942 & 0.965 & 1.100 & 1.100 & -0.276 & -0.289 \\ 
    25 & 0.005 & 0.934 & 0.955 & 1.057 & 1.057 & -0.143 & -0.159 \\ 
    50 & -0.002 & 0.929 & 0.959 & 1.027 & 1.026 & -0.053 & -0.075 \\ 
   100 & 0.000 & 0.920 & 0.946 & 1.013 & 1.013 & 0.022 & -0.017 \\ 
   \hline
\end{tabular}
\end{table}

@@

\textbf{$\widehat{R}$}. Steep drop down up-to iteration twenty, then more gradual decrease upto iteration fourty, more or less stable after that. Conventional threshold $\widehat{R} < 1.2$ reached after seven or eight iterations, $\widehat{R} < 1.1$ after fourteen iterations for chain means and sixteen for chain variances, $\widehat{R} < 1.01$ not reached within the 100 iterations (see online appendix for full table of results).

\textbf{Auto-correlation.} Within the 100 iterations considered in this study, AC decreases steeply, then slowly increases. Lowest AC observed is -0.735 at four iterations. There appears to be a trend towards zero from four iterations upwards. Auto-correlation does not plateau completely. The average absolute maximum ACs are negative up to 77 iterations for chain means, and for chain variances all iterations considered here have negative ACs.

% \textbf{[Add simulation time (computational cost) to table? Add empirical SE of all diagnostics? Sample some random simulations and check distribution and plausibility of imputations? Use chain variance as scalar of interest for rhat and autocorrelation?]}

%% --------------------------------
\subsection{Simulation diagnostics}
\textbf{Bias.} From two iterations upwards, average bias across repititions is stable. Across iterations two to one hundred, bias fluctuates within a narrow range around zero. 
% Average bias is not zero because there is only one incomplete dataset to be imputed. Bias is a sample effect.

\textbf{Confidence interval width.} CIW does not show a clear trend across iterations. One might argue that there is some downward trending upto fourty iterations. %After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached 0.928.

\textbf{Coverage rate.} Coverage rate is more or less stable from two iterations upwards. On average, the coverage rate is somewhat higher then the expected nominal coverage of 95\% \cite{neym43}, namely 0.953.

@@

\textbf{Summary.}
This note illustrates that conventional convergence diagnostics behave differently on MI data than other MCMC methods like Gibbs samplers in Bayesian analyses. The most recent recommended threshold for $\widehat{R}$ ($\widehat{R} < 1.01$) is too stringent for MI data.

From the simulation diagnostics it appears that as little as two iterations could be sufficient to draw valid inference. Convergence diagnostics $\widehat{R}$ and auto-correlation, however, indicate deviant behavior upto twenty or even fourty iterations.

\textbf{R hat below 1.}
Moreover, $\widehat{R}$ could theoretically not be smaller than one, yet it happened several times in this study (see online appendix XYZ). How could $\widehat{R}$ smaller than 1 occur? The number of simulations is smaller than in `regular' MCMC processes [explain that less iterations is an advantage of MI, not a disadvantage compared to MCMC]. Therefore, the `$(n-1/n)$' [add equation number] correction factor can influence the estimated potential scale reduction factor. This downwards bias is in the opposite direction than expected: ``The mixture-of-sequences variance, V, should stabilize as a function of n. (Before convergence, we expect $\sigma^2$ to decrease with n, only increasing if the sequences explore a new area of parameter space, which would imply that the original sequences were not overdispersed for the particular scalar summary being monitored.)'' \cite[p~438]{broo98}.

@@

\textbf{Negative ACs.} Auto-correlation is dangerous when positive. These auto-correlations are negative. Still, we want the iterations to be stable and independent. Default maxit is five iterations now, should this be different? Are the 'waves' most pronounced at iteration 5? But why the dip??

Implication of dip in AC is that default maxit value of five iterations is the worst possible number of iterations.

\textbf{Future research.} We would like to have convergence measures for multivariable statistics (scalars?) of interest. This is, however, dependent of the complete data model. The eigenvector decomposition method proposed by McKay (?) should be implemented. I could not find any resources to apply this method and it is outside the scope of this thesis to investigate how this approach could be implemented. \textbf{[this paragraph is too negative, explain scope in intro instead.]}
